Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-05-08 msc-publication.tex
fix streamfunction, add acknowledgements

\title{Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model}
\authors{James Shaw\correspondingauthor{Department of Meteorology, University of Reading, Earley Gate, PO Box 243, Reading, RG6 6BB, UK.}}
\affiliation{}
\email{j.shaw@pgr.reading.ac.uk}

@@

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain envelope half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

@@

To ensure the discrete velocity field is non-divergent, velocities are prescribed at cell faces by summing the streamfunction, \(\Psi\), at the centre of all face edges.  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain using a centred linear scheme and the upwind-biased cubic scheme.  Tracer contours at \(t = \SI{0}{\second}, \SI{5000}{\second}\) and \(\SI{10000}{\second}\) are shown in Figure~\ref{fig:advection-tracer}.  We do not compare with the results from \citet{schaer2002} because we were unable to reproduce their result using a centred linear scheme (their figure 6a), though similar results were obtained by halving the mountain height, \(h_0\).

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2}, \nonumber \\
	\frac{\partial h}{\partial x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{1.0em}} l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{0.0318} & \num{-0.442} & \num{0.849} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00251} & \num{-0.0244} & \num{0.985} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{0.00233} & \num{-0.0253} & \num{0.945} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0298} & \num{-0.258} & \num{0.865} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0318} & \num{-0.270} & \num{0.846} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No terrain & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & {---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

\acknowledgments
\TODO{John, Terry. Assuming only Hilary is co-author...?} Weller is funded by NERC grant NE/H015698/1.

>>>>

\title{Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model}
\authors{James Shaw\correspondingauthor{Department of Meteorology, University of Reading, Earley Gate, PO Box 243, Reading, RG6 6BB, UK.} and Hilary Weller}
\affiliation{Department of Meteorology, University of Reading, Reading, United Kingdom}
\email{js102@zepler.net}

@@

\TODO{take care to explain domain and grid because of our previous differences with Schaer's results} The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain envelope half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

@@

To ensure the discrete velocity field is non-divergent, velocities are prescribed at cell faces by summing the streamfunction, \(\Psi\), at the centre of all face edges.  Since \(u = - \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain using a centred linear scheme and the upwind-biased cubic scheme.  Tracer contours at \(t = \SI{0}{\second}, \SI{5000}{\second}\) and \(\SI{10000}{\second}\) are shown in Figure~\ref{fig:advection-tracer}.

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = - u_0 H \frac{z - h}{H - h}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= - \frac{\partial \Psi}{\partial z} = - u_0 \frac{H}{H - h}, \quad w = \frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{z - H}{\left( H - h \right)^2}, \nonumber \\
	\frac{\partial h}{\partial x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{1.0em}} l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l | l @{\hspace{1.0em}} l @{\hspace{1.0em}} l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{9999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} \\
SLEVE & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{9999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} \\
Cut cell & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{9999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} \\
No terrain & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \num{999} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & {---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

\acknowledgments
I am grateful to my cosupervisors John Methven and Terry Davies for their valuable input, and to Christoph Sch\"{a}r for his assistance in reproducing his advection test results.  I am thankful for the NERC studentship which helped fund this work.  Weller is funded by NERC grant NE/H015698/1.
